paper1_figures

### 2023-11-03 scoping_review_short.Rmd
short version of gender, social status, SVO and trust

@@

The effect of gender on cooperation is controversial, and studies that investigated it provided contradictory results. Sometimes men cooperated more [@Dorrough2019], sometimes women did [@Peshkovskaya2017], but, in general, meta-analysis found a non significant effect size [e.g. @Balliet2011; @Spadaro2023]. This indicates that gender differences in cooperation are very flexible and can vary depending on the context [@Spadaro2023]. Therefore, when a significant effect is observed, it means that gender has somewhat been made salient by the situation [@Sell2009].

Interaction partners can moderate this effect. In fact, in a social dilemma experiment, women were more cooperative toward strangers than men, while men cooperated more in a group of friends. Moreover, after a socialization phase where strangers could interact, men cooperated more than women [@Peshkovskaya2017]. These results seem to indicate that men accord more importance at the relationship with other members than women. In fact, the creation of a team identity enhances men's, but not women's, cooperation [@Gomez-Ruiz2020]. Moreover, in a social dilemma with many iterations, men, but not women, tended to increase their cooperation over time [@Balliet2011].

The partners' gender also have an impact on women's and men's cooperation. In general, groups composed by men only are more cooperative than only-women groups. This is explained by the fact that women, compared to men, are generally less concerned by the maximization of their personal payoff and more about relative payoff. Thus, the defection strategy allows them to avoid earning less than the other players [@Colman2018]. Moreover, men display higher levels of gratitude and trust in same-sex groups than women [@Peshkovskaya2019]. On the other side, women cooperate more in mixed group. In fact, when both men and women are present, gender stereotypes are activated and people tend to adapt their behaviour to them. This leads women to behave in a more pro-social, altruistic way and men to be more self-oriented and competitive [@Balliet2011]. The same result is also found by @Peshkovskaya2019, who additionally demonstrated that women reported a stronger social connection with their partners in mixed than in same-sex groups. The overall group composition also have an impact on cooperation: in general, groups with a majority of women display the highest levels of cooperation [@Spadaro2023]. However, @Barrero-Amortegui2021 showed that groups with a majority of women were more cooperative only after the signature of an agreement against excessive harvesting of a common resource. Therefore, even the effect of group composition seems to be subjected to moderator variables.

Another reason why gender differences in cooperation are often very small is that fear and greed motives are confused in a classical social dilemma situation [@Simpson2003]. In this contexts, a feeling of fear arises when one's cooperation could be exploited by others' free-riding, while greed is the feeling associated with the temptation to free-ride on others' cooperation. According to social psychology and evolutionary theories, the author expected women to defect more when fear incentives were present, and men to defect more with greed incentives. Although only the second hypothesis was corroborated, his intuition about women being more sensitive to the fear of being exploited is confirmed by other studies. For example, women level of cooperation is predicted by their trust in their interaction partner, while men's trust and cooperation are not linked [@Irwin2015]. Moreover, women's lower cooperation is explained by the fact that they expected their partner to contribute less to the common good [@Dorrough2019].

Finally, some of the analyzed papers presented interesting secondary results. First, against their authors' expectations, some intercultural studies showed that the gender difference in cooperation is not explained by the level of gender equality of the country [@Dorrough2019; @Spadaro2023]. Secondly, in an experiment not involving a social dilemma, women were significantly more altruistic than men [@Dorrough2019]. According to the authors, this underlines the importance of distinguishing between prosociality and strategic interactions that characterize social dilemmas. Therefore, in a situation where classical attributes of social dilemmas are not explicit, gender difference in cooperation could not be subjected to the same moderator effects described above. Lastly, according to @Sell2009, it is not possible to think about gender without relying it to social status, with men representing the high-status group. Therefore, gender differences in cooperation could also be caused by social status: to have a better understanding of this hypothesis, it would be interesting to investigate the difference in cooperation between two other social groups with a different status (e.g. high and low socioeconomic status).

In conclusion, "at times women are more cooperative than men, less cooperative than men, or equally as cooperative as men. These variations are explained, in large part, by structural differences and identities triggered by those differences" [@Sell2009, p. 322]. The context in which men and women interact is therefore essential to predict if a gender will cooperate more than the other, and which one.

@@

Few studies have investigated how social status impact cooperation, and their results do not allow to draw a clear conclusion about this effect. For instance, advantaged members (i.e. high-status class) tend to cooperate less than disadvantaged ones [@Aksoy2019]. In fact, when participants did not receive the same endowment at the beginning of the experiment, contributions of "rich" and "poor" members converged [@Kingsley2016]. In other words, proportionally to their resources, low-status participants cooperated more. However, in their review, @VanLange2013 identified some studies that observed the opposite effect, namely that wealthier people (or those who benefited more from a common resource) contributed more to the common good than the other.

Social status can also create inter-class conflicts. In fact, people cooperated more with the members of the same social class than with other classes' members [@Aksoy2019]. Moreover, a form of solidarity is observed between low-status members: in a study where participants had the opportunity to punish others for a non-cooperative behaviour, low-status members punished significantly less other low-status participants [@Chen2022]. Also, in another experiment, participants were randomly advantaged during the interaction: subjects who received a lot of advantages were discriminated by others, and especially by the ones who were rarely advantaged [@Camera2020].

@@

In conclusion, social status impacts cooperation, but it seems to have multiple effects. Moreover, similarly to what @Sell2009 affirmed about the effects of gender and social status, @Aksoy2019 demonstrated that the lower level of cooperation between members of different social classes is mainly due to actors' identification with their classes. This shows that social identity and social status are deeply connected, and that it is difficult to separate their effects on cooperation.

@@

Social identity have a consensual positive effect on cooperation: the more individuals feel a strong identification with a group, the more they cooperate with other members of the group [e.g. @Arora2016; @Noonan2016; @Gomez-Ruiz2020]. For example, in a field study about community-organized dog parks in the United States, @Noonan2016 found that a strong sense of community was related to a higher amount of voluntary activities to maintain the parks in good conditions. This is explained by the fact that, when people strongly identify with a social group, they are more concerned about their group's well-being and reputation [@VanVugt2009] and they accord more importance to group-related goal [@Arora2016]. A strong social identity also maintains a high level of cooperation over time [@Zhang2019]. Moreover, the link between group identification and cooperation is the same for individualist and collectivist people [@Chen2007], which is a further confirmation of the strength of this effect. Group identification can also be easily implemented in a laboratory experiment, and it has still a strong positive effect on cooperation [@Wit1992]. Finally, independently of the strength of group identification, people cooperate more with other people in the same group than with out-group people [@Dorrough2015; @Aksoy2019].

@@

Personal values and personality traits can play an important role in predicting cooperative behaviour [@Hilbig2018]. One of the most studied is the Social Value Orientation (SVO), that divides people in pro-social and pro-self based on how much they value their own and others' outcomes in situations of interdependence [@Balliet2009]. The effect of SVO on cooperation is very consensual: in a meta-analysis, 81 out of the 82 papers examined reported that pro-social cooperated significantly more than pro-self [@Balliet2009].

However, this general effect is subjected to several moderators. For example, pro-self individuals cooperated more in a game where economic incentives to cooperation were present than in a game without such incentives, while pro-social did not display this difference and showed high levels of cooperation in both contexts [@Emonds2011]. A similar effect is also described by @Balliet2009, who found a stronger effect of SVO in unpaid games than in paid ones, because, when monetary rewards were present, pro-self increased their cooperation. Therefore, pro-self individuals also cooperate, but only when cooperation serves their personal interest. In fact, in a study about commuting habits, the extent to which the cooperative behaviour (i.e. choosing to commute by public transportation or carpooling) was perceived as efficient impacted pro-self individual's decision more than pro-social's [@VanLange1998].

Moreover, SVO has a stronger effect if there is an uncertainty about the number of individuals having access to a common resource: when the number of other participants was not known, pro-social, but not pro-self, reduced their harvesting comparing to a group-size certainty condition [@DeKwaadsteniet2008]. The same authors demonstrated that this result is explained by the fact that pro-social individuals want to maximize the joint outcome, while pro-self mainly think at their own personal interest. Also, SVO effect is partly explained by trust, i.e. the extent to which people think that their interaction partners will also cooperate. In fact, in @VanLange1998, pro-social individuals with a low level of trust did not make different decisions than pro-self individuals, while the most cooperative were pro-social with an high level of trust. Finally, the intuitive choice is to cooperate for pro-social individuals, while it is to defect for pro-self [@Lu2019]. However, pro-self can cooperate more if they have the possibility to deliberate about the situation, because they realize that the best strategy to maximize the joint (and thus personal) outcome is cooperation.

Other studies also focused on how much people are self- or social-oriented by using other constructs. For example, @Liu2009a demonstrated that priming a common fate to induce a context of interdependence significantly increased the cooperation level of participants with an independent self, but had no effect on participants with an interdependent self (who already displayed high levels of cooperation). Moreover, in a social dilemma experiment, competitive, individualistic people were motivated just by personal gain, while people who saw themselves as a part of a group were more interested on maximizing the income of each member, even at the cost to have a smaller personal payoff [@Probst1999]. Also, high allocentric (i.e collectivist) individuals cooperated more than low allocentric ones when the perceived social norms are cooperative: this means that they are more sensitive to social cues and therefore they adapt their behaviour to social norms [@Chen2007].

Finally, there are other personal values and traits that can affect cooperation. For example, pro-environmental values are negatively correlated with harvesting from a common resource, but only when the resource is scarce; therefore these values are more likely to motivate difficult, rather than easy, ecological behaviours [@Sussman2016]. Levels of harvesting from a common resource are also lower between left-wingers than right-wingers [@Fosgaard2019]. Moreover, humility and honesty are significant predictors of cooperation [@Kieslich2014], and this is due to the fact that people with such characteristics manage to resist to the temptation to exploit others' contributions [@Hilbig2018]. Self-control also plays a role, in the sense that more self-control leads to a greater cooperation [@Kocher2017]. Lastly, people with an higher justice sensitivity make bigger contributions to a public good [@Schlosser2018].

In conclusion, values and personality traits have a large impact on human cooperation, and it is thus very important to take them into account when investigating this phenomenon. Interestingly, some social psychological techniques (e.g. elaboration) allow to modify an individual's values [@Sussman2016]. Such techniques could therefore be used to improve cooperation.

@@

Trust can be defined as the "expectations of others' benevolent motives in situations that involve a conflict between self and collective interests" [@Balliet2013, p.1]. Trust can be dispositional or a state; the former is an individual's general trait, and is usually measured with personality scales, the latter is linked to the context, and can be operazionalized by expectations about a partner's behaviour in a specific situation [@Balliet2013]. The two are positively correlated [@Lubke2021], and both have a positive general effect on cooperation [@VanLange2013; @Balliet2013]. In fact, high levels of trust are associated with more contributions to a public good [@Kocher2015], and a participant's contribution in a public good game can be predicted by her beliefs about others' contributions [@Smith2013].

Several studied confirmed this relationship. For instance, @Chaudhuri2002 demonstrated that people who cooperated more in a first public good game trusted more their partner in a following interaction than people whose main strategy was defecting. Moreover, people that were pessimistic about others' contribution cooperated significantly less than the optimistic [@Oyediran2018], and those who free-rode in a public good game, compared to cooperators, believed that others contributed less [@Kocher2015]. Also, people were more willing to purchase actions for a CO2 mitigation initiative when they believed that the majority of others had also bought them [@Sturm2019]. Finally, in a study about self-administered outlets in Swiss farms (i.e. outlets where nobody monitors if people pay or not their purchases), @Franzen2019 found that a farmer's general trust in other people is a significant predictor of the probability of having such an outlet. Concerning the moderators, trust matters more in situations involving a large, rather than small, conflict between personal and collective interests [@Balliet2013], and when people are uncertain about others' intentions [@VanLange2013].

People can also display different levels of trust in the "reliability, effectiveness and legitimacy of public institutions" [@Rompf2017, p. 2]. These authors showed that institutional trust is associated with the probability of recycling, because knowing that sanctions can be efficiently applied can motivate people to comply to norms. Moreover, the effects of perceived costs and benefits of the action are less predictive in people with high institutional trust.

Finally, trusting others (and the institutions) is essential to display higher level of cooperation. However, general beliefs about others' cooperation are quite low compared to the actual behaviours, and this trust gap can undermine cooperation, since people's actions are significantly predicted by their beliefs [@Lubke2021]. In order to increase trust and, consequently, cooperation, it is therefore essential to provide people with reliable information about others' behaviour.

>>>>

Five individuals factors that can impact cooperation emerged from the litterature review: gender, social status, group identification, values and personality traits, and trust. In the following sections, the principal effects of these factors on cooperation are described. Moreover, an overview of the moderators of each effect is provided. More details about moderators are included in the Annexe A.

@@

The effect of gender on cooperation is controversial, and studies that investigated it provided contradictory results. Sometimes men cooperated more [@Dorrough2019], sometimes women did [@Peshkovskaya2017], but two meta-analysis found a non significant effect size [@Balliet2011; @Spadaro2023]. This indicates that gender differences in cooperation are very flexible and depend on the context [@Spadaro2023]. More specifically, @Sell2009 explained this contradictory results "by structural differences [of the experimental settings] and identities triggered by those differences" (p. 322).

The effect of gender on cooperation is therefore subjected to several moderators, since the situation in which women and men interact is essential to determine which gender will cooperate more. For instance, men are more cooperative when they know the other participants [@Peshkovskaya2017], when they feel part of a team [@Gomez-Ruiz2020] or when they interact repeatedly with the same group [@Balliet2011]. The partners' gender is also an important moderator. Men-only groups are more cooperative than women-only [@Colman2018; @Peshkovskaya2019], while women cooperate more in mixed groups [@Balliet2011; @Peshkovskaya2019]. In general, groups with a majority of women display the highest levels of cooperation [@Spadaro2023], even if this effect of group composition can be moderated by other factos [e.g. @Barrero-Amortegui2021]. Moreover, men are more sensitive to greed incentives than women , and thus they cooperate less in situations with strong temptations to free-ride are present [@Simpson2003]. On the other hand, women's principal motive in cooperative interaction is avoiding to be exploited, and that is why the cooperate less if they don't trust other participants or if they believe than others provide small contributions [@Irwin2015; @Dorrough2019].

Lastly, gender differences are also linked to social status, with women representing the low-status group. Therefore, gender differences in cooperation could also be partly explained by social status: to have a better understanding of this hypothesis, it is important to understand how social status impact cooperation.

@@

The results of studies investigating the impact of social status on cooperation do not allow to draw a clear conclusion about this effect. For instance, some findings suggest that low-status participants cooperate more than high-status ones by contributing a higher proportion of their endowments [@Kingsley2016]. However, @VanLange2013 identified some studies where wealthier people cooperated more.

Social status can also create inter-class conflicts, because people tend to cooperate more with the members of the same social class than with other classes' members [@Aksoy2019]. This is especially true for low-status groups, where solidarity can arise either in the form of discrimination toward high-status members [@Camera2020], or of less punishments for other low-status [@Chen2022].

@@

Lastly, similarly to what @Sell2009 affirmed about the effects of gender and social status, @Aksoy2019 demonstrated that the lower level of cooperation between members of different social classes is mainly due to actors' identification with their classes. This shows that social identity and social status are deeply connected, and that it is difficult to separate their effects on cooperation.

@@

Social identity have a widely demonstrated positive effect on cooperation: the more individuals identify with a group, the more they cooperate with other members of the group [e.g. @Arora2016; @Noonan2016; @Gomez-Ruiz2020]. In fact, when people strongly identify with a social group, they are more concerned about their group's well-being and reputation [@VanVugt2009] and they accord more importance to group-related goals [@Arora2016]. A strong social identity also maintains a high level of cooperation over time [@Zhang2019]. The strength of this effect is further confirmed by the fact that it is the same for individualist and collectivist individuals [@Chen2007], and that it works even when groups are created in a laboratory experiment [@Wit1992].

Moreover, the link between group identification and cooperation is the same for individualist and collectivist people [@Chen2007], which is a further confirmation of the strength of this effect. Group identification can also be easily implemented in a laboratory experiment, and it has still a strong positive effect on cooperation [@Wit1992]. Finally, independently of the strength of group identification, people cooperate more with other people in the same group than with out-group people [@Dorrough2015; @Aksoy2019].

@@

Personal values and personality traits play an important role in predicting cooperative behaviour [@Hilbig2018]. One of the most studied is the Social Value Orientation (SVO), that divides people in pro-social and pro-self based on how much they value their own and others' outcomes in situations of interdependence [@Balliet2009]. The effect of SVO on cooperation has been widely demonstrated: in a meta-analysis, 81 out of the 82 papers examined reported that pro-social cooperated significantly more than pro-self [@Balliet2009].

There are several moderators to this main effect. For instance, when economic incentives to cooperation are present, pro-self individuals attain a level of cooperation similar to pro-social's one [@Emonds2011; @Balliet2009]. Similarly, in a study about commuting habits, the extent to which the cooperative behaviour (i.e. choosing to commute by public transportation or carpooling) was perceived as efficient impacted pro-self individual's decision more than pro-social's [@VanLange1998]. Another moderator is the uncertainity about how many individuals compose the group: if the uncerainty is high pro-social, but not pro-self, increse their level of cooperation [@DeKwaadsteniet2008]. Finally, pro-self increase their cooperation if they have the possibility to deliberate about the situation [@Lu2019].

Other studies also focused on how much people are self- or social-orientation impact cooperation by using other constructs. More cooperative individuals have an interdependent definition of the self [@Liu2009a] and see themselves as part of the group [@Probst1999]. Moreover, allocentric (i.e. collectivist) people are more sensitive to cooperative social norms [@Chen2007]. Finally, other personal values and traits that affect cooperation include pro-environmental values [@Sussman2016], political orientation [@Fosgaard2019], humility and honesty [@Kieslich2014; Hilbig2018], self-control [@Kocher], and justice sensitivity [@Schlosser2018].

@@

Trust can be defined as the "expectations of others' benevolent motives in situations that involve a conflict between self and collective interests" [@Balliet2013, p.1]. There are two category of trust: first, dispositional trust is an individual general trait that can be assessed with personality scale. Secondly, state trust is linked to the context and represent the expectations about a pertner's behaviour in a specific situation [@Balliet2013]. The two are positively correlated [@Lubke2021], and both have a positive general effect on cooperation [@VanLange2013; @Balliet2013].

Several studied confirmed this relationship: high levels of trust are associated with more contributions to a public good [@Kocher2015], individuals' cooperation can be predicted by their beliefs about others' contributions [@Smith2013], people who are pessimistic about others' contribution cooperate significantly less than optimistic [@Oyediran2018], and individuals who cooperate more display higher levels of trust in following interactions [@Chaudhuri2002]. The effect of trust on cooperation has also been demostrated in real-life studies [e.g. @Sturm2019; @Franzen2019]. Finally, trust matters more in situations involving a large, rather than small, conflict between personal and collective interests [@Balliet2013], and when people are uncertain about others' intentions [@VanLange2013].

People can also display different levels of trust in the "reliability, effectiveness and legitimacy of public institutions" [@Rompf2017, p. 2]. Institutional trust is associated with the probability of recycling, because knowing that sanctions can be efficiently applied can motivate people to comply to norms. Moreover, the effects of perceived costs and benefits of the action are less predictive in people with high institutional trust.
